Contributions of adaptation and purifying selection to SARS-CoV-2 evolution

### 2022-10-30 manuscript/response_to_reviewers.tex

\subsection*{Dear Drs. Stern and Duffy,}

\section*{Editor}

\editor{

Three reviewers have seen the paper, and I completely agree with all three that this is an excellent paper that uses a robust \& rapid approach to obtain meaningful insights about SARS-CoV-2 evolution. The paper could be published in its current form but would benefit from very minor corrections suggested by the reviewers.
I also add a very small concern of my own: citing from the paper "...remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations". This is a little ad-hoc, how much would the results change if changing this? How many sequences are removed this way? To emphasize, I don't think any in-depth analysis is necessary here, just a few sentences on why this cutoff is used.
Looking forward to seeing this paper published,
Adi
}

@@

\response{Thanks a lot for handling the paper, the constructive comments, and the rapid turn-around.

You are completely right that these cutoffs are heuristic and I have checked that the results are robust.

The number of outliers removed varies from clade to clade as the clades, but never exceeds 1\% with the standard parameters.
Changing these parameters (using 4 instead of 2 sigma of the Poisson distribution, 6 instead or 3 mutations offset which reduces outlier removal by 90\%) changes the rate estimates by a less than one percent (3\% at most for clade 20H).
The relative insensitivity to these outliers is due to the fact that we don't build a tree and a few outliers don't distort the overall picture.
Similarly, making the QC filter less stringent (using 80 instead of 30) doesn't change rate estimate by more than 2\% on average. The biggest effect of the QC filter is filtering for completeness.
I have added a few sentences to highlight this.

Please find my response to the reviewer comments below.

best,
Richard

@@

\refb{3. What is the ordering in Fig. S1? The order in which Nextstrain clades were assigned? I assume this is correlated to the order of clade emergence, but not necessarily exactly the same, nor would I expect the distance from the reference to the clade founder to have the same ordering. Would it make more sense to draw this figure in the same way as Fig. 4?}

\response{I have played with this, but I think the explicit ordering by clade is more informative (too many nearly overlapping markers otherwise). The clades are labeled in their order of designation and thus almost in their order of designation. I have added a statement to the figure caption. }

\refb{4. Could ORF1a and ORF1b be split in Fig. S1? It is mentioned that ORF1b is under the most evolutionary constraint (Fig. 6), but when combined with ORF1a as in Fig. S1 it appears to have the fastest rate of evolution.}

\response{I think the primary issue here is that there rates are per gene, and not per site. The objective here was to show how the total non-synonymous change is broken down by gene. I have now split `ORF1ab'. The updated plot shows that most of the observed evolution is indeed in ORF1a. }

@@

\refc{I would be interested to have a definition (or a citation) of what constitutes a nextstrain clade -- I assume it is phylogenetic, but just a quick summation of what the rules are to define it would be helpful, especially since it is different to the more commonly used pango lineage system.}

\response{Yes, Nextstrain clades are phylogenetically defined and assigned using Nextclade. I have added a reference to a blog post on nextstrain.org that outlines the clade definitions.}

\refc{Paragraph starting ``In other cases, notable subclades…'' page 3 line 55 in the second column -- I don't follow the logic here of increased divergence ``generating an offset to diversity''. What is diversity if not within-clade divergence?}

\response{Yes, this was poorly worded. What I mean to say is that when there are many lineages, time since emergence can estimated accurately from the divergence of many independent evolutionary trajectories. When there is a single subclade or a few large subclades, estimates of emergence times are more noisy due to shared ancestry. I have made an effort to reword this. }

@@

\refc{What are the advantages of this method over using BEAST to estimate rates?}

\response{The primary advantages are (i) being able to use more data, (ii) being more robust to problematic or misdated sequences, and (iii) fewer cryptic dependencies on model and prior choices. Disadvantages compared to phylogenetic methods is not taking into account shared ancestry between sequences. But since this method is only applied to variation within clades that have rapidly expanded, this is a minor issue. }

### 2022-10-30 manuscript/variant_rates.tex
final edits

\begin{abstract}
Continued evolution and adaptation of SARS-CoV-2 has lead to more transmissible and immune-evasive variants with profound impact on the course of the pandemic.
Here I analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes separately for evolution within clades -- well defined mono-phyletic groups with gradual evolution -- and for the pandemic overall.
The rate of synonymous mutations is found to be around 6 changes per year.
Synonymous rates within variants vary little from variant to variant and are compatible with the overall rate.
In contrast, the rate at which variants accumulate amino acid changes (non-synonymous mutation) was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
The overall rate of non-synonymous evolution, that is across variants, is estimated to be about 25 amino acid changes per year.
This 2-fold higher rate indicates that the evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains and likely dominated by adaptive evolution.
I further quantify the spectrum of mutations and purifying selection in different SARS-CoV-2 proteins.
Many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
About half of the mutations in other proteins are strongly deleterious and rarely observed, not even at low frequency.
\end{abstract}

@@

Since its emergence in late 2019 \citep{zhu_novel_2020}, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
Many of these variants spread considerably faster and quickly displaced the resident variants either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
Specific Variants of Concern (VoC) or Interest (VoI) were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
The branches leading to these variants are characterized by many amino acid-changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.

@@

The dichotomous pattern of SARS-CoV-2 evolution with step-wise evolution within clades or variants and atypical bursts of evolution leading to new variants has been investigated by \citet{tay_emergence_2022} and \citet{hill_origins_2022}, who showed that the rate of evolution along branches giving rise to new variants is up to four-fold higher than the background rate.
Here, I build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
This comparison reveals a consistent dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
This difference in evolutionary rate is only seen for non-synonymous changes -- the rate of synonymous evolution within variants is compatible with that seen between variants.
Furthermore, early variants display more rapid non-synonymous evolution than later variants suggesting more ubiquitous adaptive evolution early on.
I further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome.

@@

Evolutionary rates and divergence times are typically estimated using phylogenetic approaches \citep{drummond_relaxed_2006}.
These methods, however, cannot handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated, recombinant, contaminated, or otherwise chimeric sequence can fatally distort the analysis, in particular if placed early in tree.
Problematic sequences can be particularly common when a new variant takes over since sequencing protocols need adjusting \citep{deMaio_issues_2020}.

To circumvent many of the above-mentioned problems and still use the majority of the available data, I use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
I first use Nextclade \citep{aksamentov_nextclade_2021} to assign 12 million sequences available in GISAID \citep{shu_gisaid_2017} to one of the Nextstrain clades \citep{hadfield_nextstrain_2018,roemer_sars-cov-2_2022}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots), see Fig.~\ref{fig:clade_overview}.
Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
These clades represent well-defined groups of sequences with little evidence of recombination within them and are analyzed independently.
In this analysis, I only consider clades that have had significant circulation for at least 6 months.

For each clade, I define a ``founder'' genotype and exclude any sequence that does not have the full set of clade defining mutations relative to the reference sequence Wuhan/Hu-1.
This founder sequence is manually curated for each clade considered.
This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence, but will ignore a few sequences with true reversion mutations.
Results are insensitive to the stringency of this filtering.
For this reduced set of sequences, I determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
The latter step is done for nucleotide changes as well as for amino acid changes.

Within each clade, the number of mutations is expected to increase linearly in time and the variation around this mean would, in an ideal case, obey Poisson statistics.
For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
To exclude these outliers, I perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations. These outliers are never more than 0.5\% of all sequences that pass the first round of filtering and the results are insensitive to the filtering criteria.

After removing these outliers, the data are binned by calendar week.
Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each time bin is weighted with the fourth root of the number of sequences in the bins.
The exact functional form of this weighting does not have a big influence on the results, but a sublinear weighting helps to counter the large variation in sequencing effort across countries and the unavoidable imbalance due to the fact that few sequences are available early after a new variant emerges.
Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
Both synonymous and non-synonymous within-clade average divergence increase linearly over time allowing for a robust estimate of the rate.

Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is not suitable to estimate evolutionary rates in general.
In particular, confidence intervals are difficult to obtain.
However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from one or several basal polytomies.
Along each of these lineages, mutation accumulation is independent.
While not every sequence is an independent sample, the effective number of independent samples is large and the steadily increasing average divergence allows to estimate the rate robustly.

A simple model for diversity within a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
The probability that offspring genomes are different from their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5$ days and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$ (note that this rate excludes strongly deleterious mutations, see Discussion).

@@

In other cases, notable subclades did become dominant during the early stochastic dynamic.
In this case, extrapolation of the linear fit to zero divergence is not necessarily a good estimator of the emergence time of a variant: if branches leading to big subclades carry anomalously many or few mutations, the divergence time will be over- or underestimated.
For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21 (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder-like variants are a minority even in early data.

@@

\begin{figure*}
    \includegraphics[width=0.9\textwidth]{figures/rate_summary.pdf}
    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B\&C show the estimated divergence of the founder genotype of each clade (big dot or square) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
    In addition, each panel contains a regression of the divergence of clade founders vs time (gray line).
    The standard deviation expected based on Poisson statistics is indicated as shaded area.
    Panel D summarizes the individual rate estimates (dots and squares) and compares them to the estimate ofinter-clade rates (gray lines).
    In panel D, clades are in alphabetical order, which is similar to their (uncertain) order of emergence.
    The red outlier is `composite' clade 19B++ with inflated rates due to adaptive mutations.
    \label{fig:rate_summary} }
\end{figure*}

Despite these caveats, for almost all Nextstrain clades, the slope at which diversity increases is robust (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C), allowing us to estimate clade-specific evolutionary rates for amino acid and synonymous changes.
These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
In addition, I estimated the rate at which the variants themselves accumulated amino acid and synonymous changes by regressing the number of changes of the variants (relative to the putative root in clade 19B \citep{caraballo-ortiz_tophap_2022}) against the estimated time of origin of the variants.
These regressions are shown in thick gray lines in Fig.~\ref{fig:rate_summary}.

Rates of synonymous change are very consistent across variants (about 5-8 changes per genome per year) and also agree with the overall rate of synonymous changes of 7.3 changes per genome per year.
The rates of non-synonymous changes are much more variable (Fig.~\ref{fig:rate_summary}D).
Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year (see Fig.~\ref{fig:rate_progression}).
In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.
The spike protein does not contribute to this decreasing trend in the rate of within-clade non-synonymous evolution, which is most evident in \emph{ORF1ab} and to a lesser degree in accessory proteins and \emph{N} (see Supp.Fig.~\ref{fig:rate_progression_by_gene}).

Nextstrain clades tend to be defined by long branches leading to a large polytomy.
It could thus be that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
This effect might be particularly important early on in the pandemic where branches with as few as two mutations were used to define new clades.
I therefore also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
The latter is due to the rapid expansion and subsequent dominance of clade 20A, the first clade with spike mutation $D614G$, and its descendants which rapidly fixed four additional mutations \citep{korber_tracking_2020}.
This is an early example of an accelerated global rate of evolution due to adaptive evolution, see Discussion.

@@

\begin{table*}
\begin{tabular}{l|rrrrrr}
    \hline
    {\bf clade} &  overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
    \hline
    19B &     19.46 &    12.37 &      7.09 &        0 &       0 &        0 \\
   19B+ &     18.56 &    11.92 &      6.64 &        0 &       0 &        0 \\
  19B++ &     27.91 &    16.41 &     11.50 &        0 &       0 &        0 \\
  19A &     19.73 &    12.79 &      6.95 &        2 &       1 &        1 \\
  20A &     15.08 &     7.94 &      7.15 &        6 &       3 &        3 \\
    20B &     14.29 &     7.69 &      6.59 &        9 &       5 &        4 \\
    20C &     16.73 &     9.92 &      6.81 &        8 &       5 &        3 \\
   20A+ &     16.30 &     9.40 &      6.90 &        6 &       3 &        3 \\
    20E &     14.04 &     5.65 &      8.39 &       13 &       5 &        8 \\
    20H (Beta) &     15.48 &     9.51 &      5.96 &       21 &      16 &        5 \\
    20I (Alpha) &     11.56 &     5.49 &      6.07 &       30 &      19 &       11 \\
    20J (Gamma) &     12.97 &     6.29 &      6.68 &       35 &      22 &       13 \\
    21D (Eta) &     11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
    21G (Lambda) &     12.04 &     6.51 &      5.53 &       28 &      19 &        9 \\
    21H (Mu) &     18.25 &    10.05 &      8.20 &       31 &      21 &       10 \\
    21I (Delta) &     16.55 &     8.93 &      7.62 &       30 &      24 &        6 \\
    21J (Delta) &     14.30 &     7.17 &      7.14 &       34 &      28 &        6 \\
    21K (Omicron) &     13.48 &     6.45 &      7.03 &       56 &      46 &       10 \\
    21L (Omicron) &      9.24 &     4.39 &      4.85 &       68 &      50 &       18 \\
    22A (Omicron) &      7.71 &     2.81 &      4.90 &       71 &      52 &       19 \\
    22B (Omicron) &     11.90 &     5.03 &      6.87 &       66 &      50 &       16 \\
           \hline
    \end{tabular}
\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.}
    The last three columns give the distances of the clade founder sequence from putative MRCA of SARS-CoV-2.
\label{tab:rates}}
\end{table*}

@@

In addition to neutral and potentially adaptive mutations, I also quantified purifying selection on the SARS-CoV-2 genome.
By analyzing the rate of mutations that spread only on short times scales within fine-grained Pango lineages, I estimated the level of constraint on different parts of the SARS-CoV-genome.
The great majority of 3rd positions in codons at which  most mutations are synonymous do not show strong signatures of conservations.
The major open reading frames \emph{ORF1ab, S, N, E} and \emph{M} show clear signatures of purifying selection with around 50\% of sites being more constrained than the most constrained 10\% of 3rd positions.
Purifying selection analyzed there is operating over the time rare lineage circulate, which is typically a few weeks.
On shorter time scales, more strongly deleterious mutations will be observed.
The degree of purifying selection is consistent with a similar analysis in HIV-1 that also show small fitness costs for most synonymous mutations, while half of the non-synonymous mutations are so deleterious that they are not even observed on short timescales \citep{zanini_vivo_2017}.

The SARS-CoV-2 \emph{ORF3a, ORF6, ORF7a, ORF7b} and \emph{ORF8} show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
This lack of strong purifying selection in these ORFs does not necessarily imply that they don't matter or are not possible loci of adaptation -- but the exact amino acid sequence does not seem essential for their function.
Only a few regions show a clear signal of conservation at 3rd positions, notably a central region of \emph{E} and the ribosome slippage region at the beginning of \emph{ORF1b}.

@@

The effects of adaptive evolution are harder to control and account for.
Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic -- it is not the typical events, but the rare and extreme events that determine the course of adaptive evolution.
Unlike neutral evolution, the rate of adaptive evolution depends on the population size, stochastic nature of the transmission process, the environment, and previous adaptation \citep{neher_genetic_2013}.
These factors make extrapolation -- whether it is reconstruction of past events or scenarios of future evolution -- uncertain.
Standard phylogenetic or phylodynamic methods will underestimate that uncertainty.

>>>>

\begin{abstract}
Continued evolution and adaptation of SARS-CoV-2 has lead to more transmissible and immune-evasive variants with profound impact on the course of the pandemic.
Here I analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes separately for evolution within clades -- well defined mono-phyletic groups with gradual evolution -- and for the pandemic overall.
The rate of synonymous mutations is found to be around 6 changes per year.
Synonymous rates within variants vary little from variant to variant and are compatible with the overall rate of 7 changes per year (or $7.5\times 10^{-4}$ per year and codon).
In contrast, the rate at which variants accumulate amino acid changes (non-synonymous mutation) was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
The overall rate of non-synonymous evolution, that is across variants, is estimated to be about 26 amino acid changes per year (or $2.7\times 10^{-3}$ per year and codon).
This strong acceleration of the overall rate compared to within clade evolution indicates that the evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains and likely dominated by adaptive evolution.
I further quantify the spectrum of mutations and purifying selection in different SARS-CoV-2 proteins and show that the massive global sampling of SARS-CoV-2 is sufficient to estimate site specific fitness costs across the entire genome.
Many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
About half of the mutations in other proteins are strongly deleterious.
\end{abstract}

@@

Since its emergence in late 2019 \citep{zhu_novel_2020}, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
Many of these variants spread considerably faster and quickly displaced the resident variants at the time either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
Specific Variants of Concern (VoC) or Interest (VoI) were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
The branches leading to these variants are characterized by many amino acid-changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.

@@

The dichotomous pattern of SARS-CoV-2 evolution with step-wise evolution within variants and atypical bursts of evolution leading to new variants has been investigated by \citet{tay_emergence_2022} and \citet{hill_origins_2022}, who showed that the rate of evolution along branches giving rise to new variants is up to four-fold higher than the background rate.
Here, I build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
This comparison reveals a consistent dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
This difference in evolutionary rate is only seen for non-synonymous changes -- the rate of synonymous evolution within variants is compatible with that seen between variants.
Furthermore, early variants display more rapid non-synonymous evolution than later variants suggesting more ubiquitous adaptive evolution early on.
I further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome from patterns of rare diversity.

@@

Evolutionary rates and divergence times are typically estimated using phylogenetic approaches \citep{drummond_relaxed_2006}.
These methods, however, cannot handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated, recombinant, contaminated, or otherwise chimeric sequence can fatally distort the analysis.
Problematic sequences can be particularly common when a new variant takes over since sequencing protocols need adjusting \citep{deMaio_issues_2020}.

To circumvent many of the above-mentioned problems and still use the majority of the available data, I use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
I first use Nextclade \citep{aksamentov_nextclade_2021} to assign 12 million sequences available in GISAID (2022-07-25) \citep{shu_gisaid_2017} to one of the Nextstrain clades \citep{hadfield_nextstrain_2018,roemer_sars-cov-2_2022}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots), see Fig.~\ref{fig:clade_overview}.
Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
These clades represent well-defined groups of sequences with little evidence of recombination within them and are analyzed independently.
In this analysis, I only consider clades that have had significant circulation for at least 6 months.

For each clade, I define a ``founder'' genotype and exclude any sequence that does not have the full set of clade defining mutations relative to the reference sequence Wuhan/Hu-1.
This founder sequence is manually curated for each clade considered.
This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence, but will ignore a few sequences with true reversion mutations.
In addition, sequences with a Nextclade QC score above 30 (80 for 21H because of an unaccounted frameshift) are removed.
Results are insensitive to the stringency of this filtering.
For this reduced set of sequences, I determine the mutations they carry on top of the founder genotype of the clade to analyze diversification and divergence within the clade.
The latter step is done for nucleotide changes as well as for amino acid changes.

Within each clade, the number of mutations is expected to increase linearly in time and the variation around this mean would, in an ideal case, obey Poisson statistics.
For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
To exclude these outliers, I perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations (see Fig.~\ref{fig:within_clade}). These outliers are never more than 1\% of all sequences that pass the first round of filtering and the results are insensitive to the filtering criteria.

After removing these outliers, the data are binned by calendar week.
Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each time bin is weighted with the fourth root of the number of sequences in the bins.
The exact functional form of this weighting does not have a big influence on the results, but a sublinear weighting helps to counter the large variation in sequencing effort across countries and the unavoidable imbalance due to the fact that few sequences are available early on when an emerging variant is still rare.
Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, a large clade that emerged in early 2020 that was common in North America and Scandinavia.
Both synonymous and non-synonymous within-clade average divergence increase linearly over time allowing for a robust estimate of the rate.

Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is general not a suitable method to estimate evolutionary rates.
In particular, confidence intervals are difficult to obtain.
However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from one or several basal polytomies.
Along each of these lineages, mutation accumulation is independent.
While not every sequence is an independent sample, the effective number of independent samples is large and the steadily increasing average divergence allows to estimate the rate robustly.

A simple model for diversity within a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
Offspring of genomes with $i$ mutations will carry $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
The probability that offspring genomes are different from their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5$ days and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$ (note that this rate excludes strongly deleterious mutations, see Discussion).

@@

In other cases, notable subclades did become dominant during the early stochastic dynamic.
In this case, extrapolation of the linear fit to zero divergence is not necessarily a good estimator of the emergence time of a variant: if branches leading to big subclades carry anomalously many or few mutations, the divergence time will be over- or underestimated.
For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21H (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder-like variants are a minority even in early data.

@@

\begin{figure*}
    \includegraphics[width=0.9\textwidth]{figures/rate_summary.pdf}
    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B\&C show the estimated divergence of the founder genotype of each clade (big dot or square) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
    In addition, each panel contains a regression of the divergence of clade founders vs time (gray line).
    The standard deviation expected based on Poisson statistics is indicated as shaded area.
    Panel D summarizes the individual rate estimates (dots and squares) and compares them to the estimate of inter-clade rates (gray lines).
    In panel D, clades are in alphabetical order, which is similar to their (uncertain) order of emergence.
    The red outlier is `composite' clade 19B++ (containing 19B, 19A, 20A, 20B, 20C) with inflated rates due to adaptive mutations on the branch leading to clade 20A.
    \label{fig:rate_summary} }
\end{figure*}

Despite these caveats, for almost all Nextstrain clades, the slope at which diversity increases is robust (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C), allowing us to estimate clade-specific evolutionary rates for amino acid and synonymous changes.
These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
In addition to these within-clade rates, I estimated the rate at which the clades themselves accumulated amino acid and synonymous changes by regressing the number of differences of the clade's founder sequence (relative to the putative root in clade 19B \citep{caraballo-ortiz_tophap_2022}) against the estimated time of origin of the clade.
These regressions are shown in thick gray lines in Fig.~\ref{fig:rate_summary}.

Rates of synonymous change are very consistent across clades (about 5-8 changes per genome per year) and also agree with the overall rate of synonymous changes of 7.3 changes per genome per year.
The rates of non-synonymous changes are much more variable (Fig.~\ref{fig:rate_summary}D).
Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year (see Fig.~\ref{fig:rate_progression}).
In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.
The spike protein does not contribute to this decreasing trend in the rate of within-clade non-synonymous evolution, which is most evident in \emph{ORF1ab} and to a lesser degree in accessory proteins and \emph{N} (see Supp.Fig.~\ref{fig:rate_progression_by_gene}).

Nextstrain clades tend to be defined by long branches leading to a large polytomy.
It could thus be that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
This effect might be particularly important early on in the pandemic when diversity was low and when branches with as few as two mutations were used to define new clades.
I therefore also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
The latter is due to the rapid expansion and subsequent dominance of clade 20A, the first clade with spike mutation $D614G$, and its descendants which rapidly fixed four additional mutations \citep{korber_tracking_2020}.
This is an early example of an accelerated global rate of evolution due to adaptive evolution, see Discussion.

@@

\begin{table*}
\begin{tabular}{l|rrrrrrr}
    \hline
    {\bf clade} & \# of seq. & overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
    \hline
    19B &           8187 & 19.46 &    12.37 &      7.09 &        0 &       0 &        0 \\
   19B+ &           22745 & 18.56 &    11.92 &      6.64 &        0 &       0 &        0 \\
  19B++ &           126823 & 27.91 &    16.41 &     11.50 &        0 &       0 &        0 \\
  19A &             14382 & 19.73 &    12.79 &      6.95 &        2 &       1 &        1 \\
  20A &             44629 & 15.08 &     7.94 &      7.15 &        6 &       3 &        3 \\
    20B &           42077 & 14.29 &     7.69 &      6.59 &        9 &       5 &        4 \\
    20C &           47865 & 16.73 &     9.92 &      6.81 &        8 &       5 &        3 \\
   20A+ &           136863 & 16.30 &     9.40 &      6.90 &        6 &       3 &        3 \\
    20E &           82240 & 14.04 &     5.65 &      8.39 &       13 &       5 &        8 \\
    20H (Beta) &    5077 & 15.48 &     9.51 &      5.96 &       21 &      16 &        5 \\
    20I (Alpha) &   277226 & 11.56 &     5.49 &      6.07 &       30 &      19 &       11 \\
    20J (Gamma) &   24970 & 12.97 &     6.29 &      6.68 &       35 &      22 &       13 \\
    21D (Eta) &     3653 & 11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
    21G (Lambda) &  4567 & 12.04 &     6.51 &      5.53 &       28 &      19 &        9 \\
    21H (Mu) &      2897 & 18.25 &    10.05 &      8.20 &       31 &      21 &       10 \\
    21I (Delta) &   86585 & 16.55 &     8.93 &      7.62 &       30 &      24 &        6 \\
    21J (Delta) &   733825 & 14.30 &     7.17 &      7.14 &       34 &      28 &        6 \\
    21K (Omicron) & 956696 & 13.48 &     6.45 &      7.03 &       56 &      46 &       10 \\
    21L (Omicron) & 690177 & 9.24 &     4.39 &      4.85 &       68 &      50 &       18 \\
    22A (Omicron) & 33451 & 7.71 &     2.81 &      4.90 &       71 &      52 &       19 \\
    22B (Omicron) & 124239 & 11.90 &     5.03 &      6.87 &       66 &      50 &       16 \\
           \hline
    \end{tabular}
\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.}
The column `\# of seq.' shows the number of sequences that entered the analysis after filtering and restriction to the first 6 months after the emergence of the variant.
    The last three columns give the distances of the clade founder sequence from putative MRCA of SARS-CoV-2 (19B).
\label{tab:rates}}
\end{table*}

@@

In addition to neutral and potentially adaptive mutations, I also quantified purifying selection on the SARS-CoV-2 genome.
By analyzing the rate of mutations that spread only on short times scales within fine-grained Pango lineages, I estimated the level of constraint on different parts of the SARS-CoV-genome.
The great majority of 3rd positions in codons -- at which most mutations are synonymous -- do not show strong signatures of conservation.
The major open reading frames \emph{ORF1ab, S, N, E} and \emph{M} show clear signatures of purifying selection with around 50\% of sites being more constrained than the most constrained 10\% of 3rd positions.
Purifying selection analyzed here is operating over the time during which rare lineage circulate, which is typically a few weeks.
On shorter time scales, more strongly deleterious mutations will be observed.
On longer time scales, weaker deleterious effects can be quantified.
The degree of purifying selection is consistent with a similar analysis in HIV-1 that also inferred small fitness costs for most synonymous mutations, while half of the non-synonymous mutations are so deleterious that they are not even observed on short timescales \citep{zanini_vivo_2017}.

The SARS-CoV-2 genes \emph{ORF3a, ORF6, ORF7a, ORF7b} and \emph{ORF8} show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
This lack of strong purifying selection in these ORFs does not necessarily imply that they don't matter or are not possible loci of adaptation -- but the exact amino acid sequence does not seem essential for their function.
Only a few regions show a clear signal of conservation at 3rd positions, notably a central region of \emph{E} and the ribosome slippage region at the beginning of \emph{ORF1b}.

@@

The effects of adaptive evolution are harder to control and account for.
Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic -- it is not the typical events, but the rare and extreme events that determine the course of adaptive evolution.
Unlike neutral evolution, the rate of adaptive evolution depends on the population size, stochastic nature of the transmission process, the environment, and previous adaptation \citep{neher_genetic_2013}.
These factors make extrapolation -- whether it is reconstruction of past events or scenarios of future evolution -- uncertain.
Standard phylogenetic or phylodynamic methods will underestimate this uncertainty.
